\documentclass{report}

\usepackage{array} 
\usepackage{comment}
\usepackage{graphicx}
\usepackage[T1]{fontenc}
\usepackage{csquotes}
\usepackage{balance}
\usepackage{setspace}
\usepackage[table]{xcolor}

\usepackage{lstautogobble}
\usepackage{subcaption}

\usepackage{standalone}
\usepackage{algorithmicx}

\usepackage{amsmath}
\usepackage{amssymb}

\usepackage{tikz}
\newcommand\setrow[6]{
  \setcounter{col}{1}
  \foreach \n in {#1, #2, #3, #4, #5, #6} {
    \edef\x{\value{col} - 0.5}
    \edef\y{6.5 - \value{row}}
    \node[anchor=center] at (\x, \y) {\n};
    \stepcounter{col}
  }
  \stepcounter{row}
}
\newcounter{row}
\newcounter{col}

\usepackage{lineno}
\linenumbers

\usepackage[normalem]{ulem}

\usepackage{hyperref}
\hypersetup{hidelinks}

\usepackage{listings}
\lstset{ %
language=C++,                % choose the language of the code
basicstyle=\ttfamily\footnotesize,       % the size of the fonts that are used for the code
commentstyle = \color{ForestGreen},
columns=fullflexible,
numbers=left,                   % where to put the line-numbers
numberstyle=\footnotesize,      % the size of the fonts that are used for the line-numbers
stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
numbersep=5pt,                  % how far the line-numbers are from the code
%backgroundcolor=\color{codeBG3},  % choose the background color. You must add \usepackage{color}
showspaces=false,               % show spaces adding particular underscores
showstringspaces=false,         % underline spaces within strings
showtabs=false,                 % show tabs within strings adding particular underscores
frame=single,           % adds a frame around the code
tabsize=2,          % sets default tabsize to 2 spaces
captionpos=b           % sets the caption-position to bottom
breaklines=true,        % sets automatic line breaking
breakatwhitespace=false,    % sets if automatic breaks should only happen at whitespace
keywordstyle=\color{blue},       % keyword style
  %language=Octave,                 % the language of the code
  otherkeywords={SearchVar,MV,TSS,tileExpr,Search,tFunc...},           % if you want to add more keywords to the set
  numberstyle=\tiny\color{black}, % the style that is used for the line-numbers
  rulecolor=\color{black},
escapeinside={<@}{@>}
} 
\definecolor{ForestGreen}{RGB}{34,139,34}
\newcommand{\todo}[1]{{\textcolor{red}{{\tt{TODO:}}\,\,#1 }}}
\newcommand{\nc}[0]{\todo{cite}}
\newcommand{\an}[1]{{\textcolor{blue}{Author's Note: #1}}}
\newcommand{\ttt}[1]{{\texttt{#1}}}
\usepackage{xspace}
\newcommand{\FormatDecisions}[0]{{\textsc{FormatDecisions}}}
\newcommand{\su}[1]{{#1$\times$}}
\graphicspath{{./graphics/}{.}{./graphics/FormatDecisions}{./graphics/RAJALC}}

\usepackage[subtle]{savetrees}
\vbadness=18000

\begin{document}

%\chapter{Introduction}

%\include{RAJALC.tex}

%\include{FormatDecisions.tex}

\include{SparseRAJA.tex}

\chapter{Conclusion}

We conclude with three avenues for future work.
Here, we introduced support for format conversions among the space of dense, rectilinear permutation formats
\footnote{The authors have been unable to find an existing name for this space of formats. We use \enquote{rectilinear} to distinguish from orderings that do not map linear threads of elements to consecutive memory, such as Z-orders or Morton orders. We use \enquote{permutation} to distinguish from the space's subset of lexicographic and colexicographic orders.}. 
One avenue of future work would examine alternative dense formats, including Z-orders, Morton orders, tiled data formats, and SOA-AOS formats.
This would require interface modifications to allow the specification of these orderings and implementation modifications to support those formats in RAJA.
Similarly, another avenue of future work would examine extending this approach to sparse data and computations. 
While this would require more significant modifications to the RAJA library to support the specification of sparse computations, such an approach could build on the symbolic iteration space description support introduced in Section~\ref{sec:SymbolicSegment}. 

Beyond expanding the space of data formats the system can support, another direction for future research would examine support for accelerators like GPUs and FPGAs.
Compared to CPU computing, programming systems for accelerators expect the programmer to take more responsibility for managing data movement.
The \FormatDecisions{} interface could be extended to allow users to mark kernels for offloading while the underlying symbolic evaluation could be used to ensure all necessary data is accessible to the accelerator in a format that is performant for that system. 

Finally, \FormatDecisions{} could be integrated with the loop schedule transformation framework of RAJALC. 
This would enable the simultaneous optimization of both the schedule and the data layout, giving the programmer even greater control over the execution of their computations.
Such an integration presents an interesting interface design problem in how to refer to the products of schedule transformations that may combine multiple loops into one. 


\bibliographystyle{abbrv}
\bibliography{dissertation.bib}

\end{document}